Probability of causal contact between interstellar civilizations through MonteCarlo simulations

### 2019-08-22 manuscript.tex
make up de las ultimas figuras
detalles del texto

In a recent work, \citep{Balbi2018} use a statistical model to analyze
the occurrence of causal contacts between civilizations in the Galaxy.
%
The author highlights the effect of evolutionary processes when
attempting to estimate the number of communicating civilizations that
might be in causal contact with an observer located on the Earth.

\citet{cirkovic_temporal_2004} also emphasize the lack of temporal
structure in the Drake equation.

@@

Some complex stochastic processes can be efficiently modeled with the
discrete--event (DE) simulation approach.
%
Simulations are suitable tools to analyze systems that evolve with
time and involve randomness.
%
In general, simulations require less assumptions and simplifications
than theoretical approaches, and can be applied to systems where a
theoretical model can not be found.
%
A system described with the DE paradigm is characterized by a set of
actors and events, where actors interact causally through a series of
events on a time line and process these events in chronological order
\citep{ptolemaeus_system_2014, chung_simulation_2003,
simulation_ross_2012}.
%
This method is well suited for the particular case of the difusion of
intelligence signals in the Galaxy.
%
Assuming some simple hypotheses, the discrete events method can be
performed taking into account a small number of variables, which
allows to analyze the results as a function of those variables.

In what follows, we describe the experimental setup chosen to estimate
the probabilities of causal contacts and several derived quantities in
terms of three independent parameters, namely, the mean time span
between the appeareance of consecutive communicating civilizations
($\tau_a$), their mean lifetime ($\tau_s$), and the maximum distance a
signal can be detected by another civilization ($D_{max}$).
%
Intuitively, the shorter $\tau_a$ and the larger $\tau_s$ and
$D_{max}$, there is a greater probability of the existence of causal
contacts between pairs of intelligent civilizations in the Galaxy
(Table \ref{T_simu_hypotheses}).
%
The causal connection can be produced without the need of the two
civilizations being concurrent or active at the same time.
%
Each CETI will fill an annular region of the Galaxy, limited by two
surfaces.
%
The leading front, or surface of first contact (SFC) grows from the
central CETI until it reaches the distance $D_{max}$.
%
The trailing front, or surface of last contact (SLC) also grows from
the central CETI, with a delay with respect to the SFC equivalent to
the lifetime of the CETI, and produces an annular region that shrinks
until it reaches $D_{max}$, when all signal from the central CETI is
lost.
%
In the 2D simulation only the intersection of these spheres with the
plane of the galaxy is relevant, and produce the corresponding
circumferences.
%
We illustrate in the Fig.~\ref{F_growing_sphere} the schematic
representation of the growing communicating sphere, over the
space–time diagrams, where time is represented on the horizontal
direction, and space is represented in the vertical direction.
%
In the upper panel (a), the sphere is growing as the surface of first
contact has not reached the maximum distance.
%
In the middle panel (b) it has reached the maximum distance, so that
it remains at the same size.
%
After a Doomsday event (bottom panel, b), the signals can still be
observed, but the surface of last contact grows.
%
In the Fig.~\ref{F_abcd}, we show a schematic representation of two
emitters, E$_1$ and E$_2$, that reach each other at different times.
%
The time span for E$_i$ is (A$_i$ , D$_i$) , for $i = {1, 2}$.
%
Emitter $i$ can listen to emitter $j$ between C$_{ij}$ and B$_{ij}$.
%
The type and length of causal contact in both directions depend on the
distance and time lag between the awakening events, the maximum
distance that a signal can reach and the time period in which each
emitter is active.
%
In the Fig.~\ref{F_messages} we show a schematic representation of the
possible cases in which a CETI A can be in causal contact with a CETI
B which appears later.
%
The duration of the causal contact in one direction depends on several
factors, mainly $\Delta$t~A , $\Delta$t~C , and $\Delta$X.

@@

\setlength{\tabcolsep}{10pt}
\begin{table}
\centering
\begin{tabular}{cl}
\hline
   \multicolumn{2}{l}{independent variables} \\
\hline
   $\tau_{a}$ & Mean temporal separation between\\
              & consecutive awakenings \\
   $\tau_{s}$ & Mean lifetime of a CETI \\
   D$_{max}$ & Maximum reach of a message \\
\hline
\multicolumn{2}{l}{fixed variables and assumptions} \\
\hline
   & All CETIs share the same statistical properties\\
   & The process is homogeneous \\
   & The scan of the sly is fully efficient\\
   & Absence of panspermia or colonization \\
   & The GHZ is a 2--dimensional ring \\
20000ly    & Inner radius of the GHZ \\
60000ly    & Outer radius of the GHZ \\
1.e6 y    & Time span of the simulation \\

\hline
\end{tabular}
\caption{Definition of independent variables and fixed paramenters 
   that are part of the simulation.}
\label{T_simu_hypotheses}
\end{table}

@@

In our simulations we also assume the simplest situations for the
growth of ETIs.
%
For example, we discard the possibility of stellar colonization
(Walters 1980) and assume that the communication aim is performed to
all directions with the same power and the same probability, i.e.,
strongly assume isotropic communication in all cases.
%
More detailed simulations could be produced considering different 
efficiency of communication or detection methods.
%
Also, a mixed paranoid/pronoid population of intelligent civilizations
could be implemented easily.
%
However, in this case we also would complicate the experimental setup,
make the results less evident.
%
The results are independent of the nature of life (organic or
artificial).
%
The lifetime of a civilization can be caused by autodestruction or by
external factors.

\subsection{Power laws vs. exponential laws}

@@

\subsection{Discrete event process}

Given a set of parameters $(\tau_a, \tau_s, D_{max})$, the functional
forms of the statistical distributions and fixed values
for the hypotheses, a discrete
event simulation is performed by keeping track of a set of variables
that change each time a meaningful event is produced.

\setlength{\tabcolsep}{10pt}
\begin{table}
\centering
\begin{tabular}{cccc}
\hline
   fixed parameter & \multicolumn{3}{c}{adopted value}\hfill \\
\hline
   inner GHZ radius\hfill & \multicolumn{3}{c}{20000 ly}\\
   outer GHZ radius & \multicolumn{3}{c}{60000 ly}\\ \\
\hline
   variable parameter & min & max & N \\
\hline
   D$_{max}$ & 1000 ly & 20000 ly & 20\\
   $\tau_{a}$ & 1000 y & 20000 y & 20\\
   $\tau_{s}$ & 1000 y & 20000 y & 20\\
\hline
\end{tabular}
\caption{Adopted values for simulation parameters. Fixed parameters
   define the GHZ.  Variable parameters define the temporal structure
   of the process and the reach of messages.}
\label{T_range_parameters}
\end{table}

The main variables that follow the evolution of the simulation are:

\begin{itemize}
   \item Position of stars.  Sampled randomly within the GHZ
   \item Time of the awakening of each CETI
   \item Time of the vanishing of each CETI
\end{itemize}

The variables that can be deduced from the previous ones are:

\begin{itemize}
   \item Number of CETIs in casual contact with at least another CETI
      at a given time.
   \item Number of CETIs as a function of time
   \item Number of CETIs that receive a message at least one time
   \item Number of CETIs that receive a message at least one time and
      successfully deliver an answer.
   \item Number distribution of waiting time to receive a message
\end{itemize}

\subsection{Exploration of the hypothesis space}

%========================================================================
%========================================================================
%========================================================================

\subsection{Model for CETIs}

@@

%======================================================================
%======================================================================
%======================================================================

The system is updated each time an event is produced.
%
The time marks for each event (A and D for each CETI and C and B for
each causal contact) are stored as a result of the simulation.
%
Also, the list of active CETIs is obtained as a function of time.
%
Some fixed variables must be set in order to carry out the simulation,
namely:

\begin{itemize}
   \item Size of the Galactic Habitable Zone,
   \item mean lifetime of a CETI, $\lambda_1$,
   \item mean waiting time for the appeareance of another CETI,
      $\lambda_2$, and
   \item maximum distance a message can be detected.
\end{itemize}

@@

The simulation was implemented on a python code, which is publically
available\footnote{https://github.com/mlares/simu\_contact}

%Variables en el programa
%\begin{itemize}
%   \item Nact: Número de CETIs activas
%   \item Nsee: Número de contactos por cada CETI
%   \item Nccc: CETIs en contacto causal. Notar que una CETI puede tener múltiples
%contactos de otras CETIs. Para cada una de ellas, guardar:
%   \item ID: ID de la CETI emisora
%   \item ID: ID de la CETI receptora
%   \item t\_hola: tiempo del primer contacto
%   \item t\_chau: tiempo del último contacto
%\end{itemize}
%
%
%Caso 1: aparece una nueva CETI
%Modificaciones al estado del sistema:
%\begin{itemize}
%   \item Agregar la CETI “A” a la lista de CETIs activas
%   \item Sortear la duración de la actividad de comunicación de la CETI “A”
%   \item Sortear el intervalo de tiempo hasta la aparición de la próxima CETI:
%t\_WakeUp\_next
%   \item Agregar la CETI “A” al árbol
%   \item Identificar todas las CETIs que están a menos de D max de la nueva CETI, y
%actualizar la lista de tiempos
%\end{itemize}
%
%Caso 2: desaparece una CETI
%Modificaciones al estado del sistema:
%\begin{itemize}
%   \item Eliminar la CETI a la lista de CETIs activas
%\end{itemize}
%
%
%Caso 3: una nueva CETI entra en contacto causal
%Modificaciones al estado del sistema:
%\begin{itemize}
%   \item Agregar la CETI B a la lista de contactos de la CETI A
%\end{itemize}
%

@@

@@

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=0.3\textwidth]{m1d2.png}%
   \includegraphics[width=0.3\textwidth]{m1d3.png}%
   \includegraphics[width=0.3\textwidth]{m1d4.png}
   \caption{Rate of CETIs that never make contact (listening) as a
   function of $\tau_a$ and $\tau_s$, for 
   $D_{max}$=10000 (left panel),
   $D_{max}$=40000 (middle panel), and
   $D_{max}$=80000 (right panel)}
   \label{F_never_contact}
\end{figure*}

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=0.3\textwidth]{m2d2.png}%
   \includegraphics[width=0.3\textwidth]{m2d3.png}%
   \includegraphics[width=0.3\textwidth]{m2d4.png}
   \caption{Rate of CETIs that listen at the moment of the
   awakening, as a funcion of $\tau_a$ and $\tau_s$}
   \label{F_C_at_A}
\end{figure*}

@@

\begin{figure}
   %
   \centering
   %
   \includegraphics[width=0.5\textwidth]{radial.png}
   %
   \caption{Fraction of CETIs making cero, one, two or three contacts
   as a function of galactocentric distance.}
   %
   \label{F_radial_frac}
   %
\end{figure}

\ack[Acknowledgement]{
%
This work was partially supported by the Consejo Nacional de
Investigaciones Cient\'{\i}ficas y T\'ecnicas (CONICET, Argentina),
the Secretar\'{\i}a de Ciencia y Tecnolog\'{\i}a, Universidad Nacional
de C\'ordoba, Argentina, and the Universidad Cat\'olica de C\'ordoba,
Argentina.
%
This research has made use of NASA's Astrophysics Data System.
%
Plots and simulations were made with software developed by the authors
using R and python languajes, plots were postprocessed with inkscape.
%
}

>>>>

In a recent work, \citep{Balbi2018} use a statistical model to analyze
the occurrence of causal contacts between civilizations in the Galaxy.
%
The author highlights the effect of evolutionary processes when
attempting to estimate the number of communicating civilizations that
might be in causal contact with an observer located on the Earth.
%
\citet{cirkovic_temporal_2004} also emphasize the lack of temporal
structure in the Drake equation.

@@

Some complex stochastic processes can be efficiently modeled with the
discrete--event (DE) simulation approach.
%
Simulations are suitable tools to analyze systems that evolve with
time and involve randomness.
%
In general, simulations require less assumptions and simplifications
than theoretical approaches, and can be applied to systems where a
theoretical model can not be found.
%
A system described with the DE paradigm is characterized by a set of
actors and events, where actors interact causally through a series of
events on a time line and process these events in chronological order
\citep{ptolemaeus_system_2014, chung_simulation_2003,
simulation_ross_2012}.
%
In the case of study, the actors are the communicating civilizations
(CETIs), and the events are all meaningfull that produce a change in
the simulation variables, i.e., in the arrangement of CETIs and their
causal contacts.
%
Since it allows to explore several models easily, 
this method is well suited for the particular case of the difusion of
intelligence signals in the Galaxy.
%
Assuming some simple hypotheses, the discrete events method can be
performed taking into account a small number of variables, which
allow to analyze the variation of the results in the model parameter space.

In what follows, we describe the experimental setup chosen to estimate
the probabilities of causal contacts and several derived quantities in
terms of three independent parameters, namely, the mean time span
between the appeareance of consecutive communicating civilizations
($\tau_a$), their mean lifetime ($\tau_s$), and the maximum distance a
signal can be detected by another civilization ($D_{max}$).
%
Intuitively, the shorter $\tau_a$ and the larger $\tau_s$ and
$D_{max}$, there is a greater probability of the existence of causal
contacts between pairs of intelligent civilizations in the Galaxy
(Table \ref{T_simu_hypotheses}).
%
We also need to propose theoretical distribution functions for both the
distribution of lifespans and the distribution of the number of CETIs
per unit time.
%
The shapes of these distributions are discussed in Sec.~\ref{SS_PDF_shape}.
%
A causal connection can be produced without the need of the two
involved civilizations being concurrent or active at the same time.
%
This is a fundamental property of the system which arises as a
consecuence of the large spatial and temporal scales, where a message
is transmited at light speed.
%
As a consecuence of their finite lifetime,
each CETI will fill a spherical shell region of the Galaxy, limited by two
surfaces.
%
The leading front, or surface of first contact (SFC) grows from the
central CETI until it reaches the maximum distance $D_{max}$.
%
Although a CETI could be old enough to transmit a message at large
distances, the limited power of the message and the dilution that runs
as the squared distance from the source, imposes a
detectability limit.
%
Thus, the volume reached causally by a CETI is initially a growing
sphere.
%
Following the end of the civilization, there is still a region which
is covered with the emited signals.
%
The trailing front, or surface of last contact (SLC) also grows from
the central CETI, with a delay with respect to the SFC equivalent to
the lifetime of the CETI, and produces a spherical shell region.
%
Any other CETI within this region will be aware of the originating CETI,
even if it has disapeared before the time of contact.
%
This region will grow if the SFC has not reached $D_{max}$, and
will shrink otherwise, until the SLC reaches $D_{max}$.
%
A that point, all signal from the central CETI is lost.
%
In the 2D simulation only the intersection of these spheres with the
plane of the galaxy is relevant, and produce the corresponding
circles for the filled spheres or annular regions for the spherical
shells.
%
We illustrate in the Fig.~\ref{F_growing_sphere} the schematic
representation of the region causally connected to the central CETI.
%
The initially growing communicating sphere, is shown over
space--time diagrams, where time is represented on the horizontal
direction, and space is represented in the vertical direction.
%
In the upper panel (a), the sphere is growing as the surface of first
contact has not reached the maximum distance.
%
In the middle panel (b) it has reached the maximum distance, so that
it remains at the same size.
%
After a Doomsday event (bottom panel, b), the signals can still be
observed, but the surface of last contact grows.
%
In the Fig.~\ref{F_abcd}, we show a schematic representation of two
emitters, E$_1$ and E$_2$, that reach each other at different times.
%
The time span for E$_i$ is (A$_i$ , D$_i$) , for $i = {1, 2}$.
%
Emitter $i$ can listen to emitter $j$ between C$_{ij}$ and B$_{ij}$.
%
The type and length of causal contact in both directions depend on the
distance and time lag between the awakening events, the maximum
distance that a signal can reach and the time period in which each
emitter is active.
%
In the Fig.~\ref{F_messages} we show a schematic representation of the
possible cases in which a CETI A can be in causal contact with a CETI
B which appears later.
%
The duration of the causal contact in one direction depends on several
factors, mainly $\Delta$t~A , $\Delta$t~C , and $\Delta$X.

@@

\setlength{\tabcolsep}{10pt}
\begin{table*}
\centering
\begin{tabular}{clccc}
\hline
   \multicolumn{2}{l}{independent variables} &min&max&Nbins\\
\hline
   $\tau_{a}$ & Mean temporal separation between consecutive awakenings & 4000 y & 200000 y & 50\\ 
   $\tau_{s}$ & Mean lifetime of a CETI                                 & 10000 y & 500000 y& 50\\ 
   D$_{max}$ & Maximum reach of a message  &  \multicolumn{3}{l}{500, 1000, 10000, 20000, 40000, 80000 lyr}  \\
\hline
   \multicolumn{4}{l}{fixed variables and assumptions} &value \\
\hline
   & statistical properties of all CETIs &equally distributed&&\\
   & Point process for the distribution in time & homogeneous &&\\
   f$_s$ & The scan of the sky & fully efficient&&1\\
   f$_p$ & panspermia or colonization &absent&&0\\
   & shape of the Galactic Habitable Zone & 2--dimensional ring &&\\
   R$_{SFC}$   & Inner radius of the GHZ     &&&20000 ly\\
   R$_{SLC}$   & Outer radius of the GHZ     &&&60000 ly\\
   t$_{max}$ & Time span of the simulation  &&&1.e6 y\\
\hline
   \multicolumn{5}{l}{discrete events}\\
\hline
   A & Awakening: a civilization starts a SETI program  &&&\\
   B & Blackout: the end of the communication chanel stops &&&\\
   C & Contact: a new causal contact is produced &&&\\
   D & Doomsday: a civilization ends its communication capabilities&&&\\
\hline

\hline
\end{tabular}
\caption{Definition of independent variables and adopted values for 
   fixed paramenters 
   that are part of the simulation.  Variable parameters define the
   spatial and temporal structure
   of the process and the maximum reach of the messages.}
\label{T_simu_hypotheses}
\end{table*}

@@

In our simulations we also assume the simplest situations for the
growth of ETIs.
%
For example, we discard the possibility of stellar colonization
(Walters 1980) and assume that the communication aim is performed to
all directions with the same power and the same probability, i.e.,
strongly assume isotropic communication in all cases.
%
More detailed simulations could be produced considering different 
efficiency of communication or detection methods.
%
Also, a mixed paranoid/pronoid population of intelligent civilizations
could be implemented easily.
%
However, in this case we also would complicate the experimental setup,
make the results less evident.
%
The results are independent of the nature of life (organic or
artificial).
%
The lifetime of a civilization can be caused by autodestruction or by
external factors.
%}}}

\subsection{Power laws vs. exponential laws}\label{SS_PDF_shape}
%{{{

@@

\subsection{Discrete event process}
%{{{

Given a parametric model $M(\tau_a, \tau_s, D_{max})$, which includes the
functional forms of the statistical distributions and fixed values for the
hypotheses, a discrete event simulation is performed by keeping track of a set
of variables that change each time a meaningful event is produced.
%
The main variables that follow the evolution of the simulation are the positions
of stars, which are sampled randomly within the GHZ, the time of the awakening
of each CETI (A event), and the time of the vanishing of each CETI (D event).
%
The variables that can be deduced from the previous ones include the number of
CETIs in casual contact with at least another CETI at a given time, the number
of CETIs as a function of time, the number of CETIs that receive a message at
least one time, the number of CETIs that receive a message at least one time and
successfully deliver an answer and the number distribution of waiting time to
receive a message.
%
All these quantities are updated each time one of the four events (A, B, C, D)
is produced.

%\subsection{Exploration of the hypothesis space}

\subsection{Model for CETIs}
%{{{

@@

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{matrix_1.pdf}
   \caption{Rate of CETIs that never make contact (listening) as a
   function of $\tau_a$ and $\tau_s$, for 
   $D_{max}$=10000 (left panel),
   $D_{max}$=40000 (middle panel), and
   $D_{max}$=80000 (right panel)}
   \label{F_never_contact}
\end{figure*}

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{matrix_2.pdf}
   \caption{Rate of CETIs that listen at the moment of the
   awakening, as a funcion of $\tau_a$ and $\tau_s$, for
   $D_{max}$=10000 (left panel),
   $D_{max}$=40000 (middle panel), and
   $D_{max}$=80000 (right panel)}
   \label{F_C_at_A}
\end{figure*}

The system is updated each time an event is produced.
%
The time marks for each event (A and D for each CETI and C and B for each causal
contact) are stored as a result of the simulation.
%
Also, the list of active CETIs is obtained as a function of time.
%
Some fixed variables must be set in order to carry out the simulation: the size
of the Galactic Habitable Zone, the mean lifetime of a CETI, $\tau_s$, the mean
waiting time for the appeareance of another CETI, $\tau_a$, and the maximum
distance a message can be detected.

@@

The simulation was implemented on a python 3 code, which is publically
available\footnote{https://github.com/mlares/simu\_contact}

@@

\begin{figure}
   %
   \centering
   %
   \includegraphics[width=0.45\textwidth]{galactocentric_frac.pdf}
   %
   \caption{Fraction of CETIs making cero, one, two or three contacts
   as a function of galactocentric distance.}
   %
   \label{F_radial_frac}
   %
\end{figure}

@@

@@

\ack[Acknowledgement]{
%
This work was partially supported by the Consejo Nacional de
Investigaciones Cient\'{\i}ficas y T\'ecnicas (CONICET, Argentina),
the Secretar\'{\i}a de Ciencia y Tecnolog\'{\i}a, Universidad Nacional
de C\'ordoba, Argentina, and the Universidad Cat\'olica de C\'ordoba,
Argentina.
%
This research has made use of NASA's Astrophysics Data System.
%
Plots and simulations were made with software developed by the authors
using R and python languajes. Plots were postprocessed with inkscape.
%
}
